\section{Related Work}
\label{sec:related}

\textbf{Linked Data Query Processing.} The concept of executing SPARQL
queries directly over Linked Data was first introduced in \cite{hartig_executing_2009},
where link traversal is used to discover sources at runtime. Then, an optimized index was proposed to capture statistics for ranking sources~
\cite{harth_data_2010}. Other previous work
\cite{ladwig_linked_2010,sihjoin_2011} proposes methods for adaptive ranking
of sources at runtime, and a mixed execution
strategy that combines local and remote Linked Data. % In this work we adopt the approach from
% \cite{harth_data_2010} and employ a source index without any runtime
% source discovery.
% along with the push-based Symmetric Hash Join
% operator from \cite{ladwig_linked_2010,sihjoin_2011}.


% \textbf{Dynamic Programming.}

% Faster enumeration of plans: \cite{moerkotte_analysis_2006,moerkotte_dynamic_2008}

% Iterative Dynamic Programming (approx):
% \cite{kossmann_iterative_2000}

%\textbf{Federated Databases and Data Integration.} 

\textbf{Query Optimization and Processing.} There is a large amount of
database research on query optimization. The dynamic programming
solution was first proposed in \cite{selinger_access_1979} and remains a
popular approach for query optimization
\cite{moerkotte_dynamic_2008}. There is also work on approximating the DP approach to increase run-time
performance in the context of distributed query processing
\cite{kossmann_iterative_2000}.

Efficiently generating optimal DAG-shaped query plans when performing
operator sharing has been addressed in
\cite{neumann_generating_2009}. In our work we also uses operator
sharing for dealing with Linked Data sources. However, the effect of this is different in our multi-objective optimization problem, where we introduce special bounds needed for pruning. 
%. We also adopt the technique for cost
%calculation from \cite{neumann_generating_2009}. 
The efficient
execution of DAG-shaped plans was discussed in \cite{Neumann_2005},
where several approaches were proposed, including the push-based
execution that is used in our implementation. 
%(also used in previous work on
%Linked Data query processing \cite{ladwig_linked_2010,sihjoin_2011}).

\textbf{Source Selection.} The problem of selecting relevant sources
has been a topic in data integration research~\cite{levy_querying_1996}. 
%In
%this setting, sources are described not only by their content, but
%also their capabilities. Algorithms have been proposed to efficiently
%perform source selection by using the source characteristics to prune
%the search space . 
However, 
%in these
%approaches, 
source selection here is a separate step that is decoupled from query
optimization. In \cite{nie_joint_2001} the authors recognize that the decoupling of
source selection and query optimization leads to overall sub-optimal
plans and propose a solution that optimizes not only for cost but
also coverage. A (weighted) utility function is proposed to combine them 
into a single measure. 
%Then,
%classic query optimization algorithms, such as DP, can be applied. 
Finding the right utility function is generally known to be difficult, especially when many objectives have to be considered. Instead, we follow a different direction, employing multi-objective optimization to produce Pareto-optimal plans that represent different trade-offs between the objectives.

\textbf{Top-k Processing.} Top-k query processing focuses on the most important (top-k) answers to a
given query~\cite{ilyas_survey_2008}. 
%It has been studied from different angles, resulting in different concepts as
%join algorithms, query optimization and indexing methods
%\cite{ilyas_survey_2008}. 
Clearly, multi-objective optimization is different from top-k
processing in that instead of a fixed number of results, a range of
plans representing different trade-offs is computed. One of these
plans may yield the same effect as top-k processing. However, the
overall problem here is finding optimal plans, instead of computing
results.

\textbf{Skyline Queries.} The skyline operation finds the Pareto
set from a potentially large set of points and is used in conjunction
with standard relational algebra \cite{boerzsoenyi_skyline_2001}. The
main research focus has been the efficient execution of such queries
by means of specialized algorithms and indexes. Our approach is
different in that instead of computing Pareto-optimal results, it computes query plans.
%However, techniques developed for skylining be
%adopted to improve the efficiency of the multi-objective DP planner,
%e.g. for pruning to obtain the Pareto-optimal (skyline) set
%of plans.

\textbf{Multiobjective Query Optimization.} To the best of our
knowledge, \cite{papadimitriou_multiobjective_2001} is the only work
addressing multi-objective query optimization, where it is studied in
the context of Mariposa \cite{stonebraker_mariposa:_1996}, a wide-area
database. The optimizer splits the query tree into subqueries
and then obtains bids from participating sites that specify a delay
and cost for delivering the result of a subquery. The goal of the
proposed multi-objective optimizer
\cite{papadimitriou_multiobjective_2001} is to obtain the Pareto
optimal set of plans with respect to cost and delay. While dynamic
programming is employed to show that the Pareto set can be computed in
polynomial time, it is not based on the classic DP algorithm
\cite{selinger_access_1979}. The problem studied there is different
because there is only a single query operation tree and for each
operation node, the optimizer has a list of alternatives for
implementing the operation. In contrast, the classic DP algorithm does
not consider only a single query tree (i.e. a single order of
operations), but builds and optimizes plans in a bottom-up fashion and
considers all possible query trees. In this work, we extend the
classic DP algorithm to support multi-objective query optimization.



% \textbf{Data Integration.}

% Bucket algorithm: \cite{levy_querying_1996}

% MiniCon algorithm: \cite{pottinger_minicon:_2001}

% Optimization of cost and coverage (utility function):
% \cite{nie_joint_2001}

% Source selection: \cite{pomares_source_2010}

% Query Planning in the Presence of Overlapping Sources:
% \cite{bleiholder_query_2006}

% survey materialized views: \cite{halevy_answering_2001}



%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "paper"
%%% End: 
